Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-20 latex/main.tex

\subsubsection*{XAC}
https://github.com/jeromekelleher/sc2ts-paper/issues/357
Evidence of the second breakpoint reported in its designation issue relies on
the presence of a large deletion inherited from BA.2*, which is currently treated as missing data by sc2ts.
The XAC samples share a recombination node with the samples of four other 
X lineages
which do not have this deletion (Figure 2C).

@@

\subsubsection*{Non-recombinants}
\subsubsection*{XN and XAU}
https://github.com/jeromekelleher/sc2ts-paper/issues/358
https://github.com/jeromekelleher/sc2ts-paper/issues/348
Both of these putative recombinants have been defined on the basis of a RHS from BA.2 and a LHS from BA.1.
In the case of XN, the supposed BA.1 mutations are [C241T, A2832G, C3037T].
In the case of XAU, this set also includes C2470T, so [C241T, C2470T, A2832G, C3037T].
However, both BA.1 and BA.2 are now defined as also having C241T and C3037T.
In addition, a very large number of BA.2 samples also have A2832G, and three BA.2-designated samples also have C2470T.
This means that there is essentially no longer any evidence for a BA.1 parent for XN, and
only one uncertain mutation that suggests a BA.1 parent for XAU.

@@

\subsection{Analysis of the Jackson et al. recombinants}

@@

% Data derived from the notebooks/jackson_recombinants.ipynb notebook
% See https://github.com/jeromekelleher/sc2ts-paper/issues/293
Jackson et al. described four groups of recombinants (named A to D) and
four singleton recombinants.
Of the 16 samples analyzed by Jackson et al, 12 were included in the ARG;
the remaining 4 were excluded because the sequences did not pass the 
QC filtering criteria, or due to a high HMM cost. Groups A-D are summarized 
in Table~\ref{tab:jackson} and show excellent concordance in terms 
of the parental Pango lineages and breakpoint intervals. All four groups
are directly associated with strongly supported recombination events 
in the ARG in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
The singleton CAMC-CBA018 has 
the parents B.1.177+B.1.1.7 and the breakpoint interval 16,177--20,133 in the ARG, 
and is strongly supported by 14 averted mutations.
The Jackson et al.\ results are almost identical with the same parent lineages, 
and breakpoint interval of interval 11,395–21,993.
The singleton MILK-103C712 has 
the parents B.1.177.16+B.1.1.7 and a breakpoint interval of 24915--27972 
in the ARG, again strongly supported by 11 averted mutations.
The Jackson et al.\ results for this lineage has the same parents,
with the breakpoint interval 26,800–-27,974.
The singleton QEUH-1067DEF has the parents B.1.1.7+B.1.177.9 
and a breakpoint interval of 7,729--10,870 in the ARG, 
supported by 11 averted mutations.
The Jackson et al.\ results for this lineage has the same parents,
with the breakpoint interval 6,953--10,872.

>>>>

\subsubsection*{XAC}
https://github.com/jeromekelleher/sc2ts-paper/issues/357

Evidence of the second breakpoint reported in its designation issue relies on
the presence of a large deletion inherited from BA.2*,
which is currently treated as missing data by sc2ts.
The XAC samples share a recombination node with
the samples of four other Pango X lineages,
which do not have this deletion (Figure 2C).

@@

\subsubsection*{Non-recombinants}
\subsubsection*{XN and XAU}

https://github.com/jeromekelleher/sc2ts-paper/issues/358
https://github.com/jeromekelleher/sc2ts-paper/issues/348

Both of these putative recombinants have been defined
on the basis of an RHS from BA.2 and a LHS from BA.1.
In the case of XN, the supposed BA.1 mutations are [C241T, A2832G, C3037T].
In the case of XAU, this set also includes C2470T, so [C241T, C2470T, A2832G, C3037T].
However, both BA.1 and BA.2 are now defined as also having C241T and C3037T.
In addition, a very large number of BA.2 samples also have A2832G, and
three BA.2-designated samples also have C2470T.
This means that there is no longer any evidence for a BA.1 parent for XN, and
only one uncertain mutation that suggests a BA.1 parent for XAU.

@@

\subsection{Analysis of the Jackson et al. recombinants}
\label{sec:jackson_recombs}

@@

Jackson et al. described
four groups of recombinants (named A to D) and four singleton recombinants.
Of the 16 samples analyzed by Jackson et al.\
12 were included in the ARG;
the remaining 4 were excluded
because the sequences did not pass the QC filtering criteria or
because of a high HMM cost.
Groups A-D are summarized in Table~\ref{tab:jackson} and show excellent concordance
in terms of the parental Pango lineages and breakpoint intervals.
All four groups are directly associated with strongly supported recombination events
in the ARG in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
In the ARG, the singleton CAMC-CBA018 has
the parents B.1.177+B.1.1.7 and
the breakpoint interval 16,177--20,133,
and is strongly supported by 14 averted mutations.
In the Jackson et al.\ results, this singleton is
almost identical with the same parent lineages and
a breakpoint interval of 11,395–-21,993.
In the ARG, the singleton MILK-103C712 has
the parents B.1.177.16+B.1.1.7 and
a breakpoint interval of 24915--27972,
again strongly supported by 11 averted mutations.
In the Jackson et al.\ results,
this lineage has the same parents,
with the breakpoint interval 26,800–-27,974.
In the ARG, the singleton QEUH-1067DEF has
the parents B.1.1.7+B.1.177.9 and
a breakpoint interval of 7,729--10,870,
supported by 11 averted mutations.
In the Jackson et al.\ results,
this lineage has the same parents,
with the breakpoint interval 6,953--10,872.

See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_jackson\_recombinants.ipynb}
{analysis\_jackson\_recombinants.ipynb}
for details.
